---
title: "B-age calculation"
output: html_document
date: "2024-10-29"
---

# 0. Load Libraries

```{r setup, include=FALSE}
# data management
library(dplyr)
library(stringr)

# data analysis
library(Hmisc)
library(rstatix)

# data visualization
library(ggplot2)
library(kableExtra)
library(knitr)

# B-age
library(methylclock)
library(EpiDISH)

# functions 
explore_df <- function(df){
  aux <- dim(df)
  message(paste0('Dataframe has:\n\tRows: ', aux[1], '\n\tCols: ', aux[2]))
}

blue0.5 <- scales::alpha('dodgerblue', .5)
hist_fun <- function(x, data){
  hist(data[,x], main = x, xlab = '', ylab = '', freq = F, col = blue0.5)
  lines(density(data[,x], na.rm = T))
}

knitr::knit_hooks$set(
   error = function(x, options) {
     paste('\n\n<div class="alert alert-danger">',
           gsub('##', '\n', gsub('^##\ Error', '**Error**', x)),
           '</div>', sep = '\n')
   },
   warning = function(x, options) {
     paste('\n\n<div class="alert alert-warning">',
           gsub('##', '\n', gsub('^##\ Warning:', '**Warning**', x)),
           '</div>', sep = '\n')
   },
   message = function(x, options) {
     paste('\n\n<div class="alert alert-info">',
           gsub('##', '\n', x),
           '</div>', sep = '\n')
   }
)

explore_tab <- function(df, lab){
  df %>%
  kable(., 
        align = rep('c', ncol(.)), 
        caption = lab) %>%
  kable_classic_2(full_width = F, html_font = 'cambria')
}
```

  
# 1. Data import

Steps for data import:

- Import betas file from: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE280206

- Split betas file into 450K and EPIC using GEO metadata

- Create a "CpGName" variable at the beginning of each matrix (see methylclock documentation)

- Request access to phenotypic data (C-age) from: 10.5281/zenodo.13981377

```{r}
# 450K array
c_age_450K # vector with patients' c-age (only 450K patients)
betas_450K # matrix 450K betas

# EPIC array
c_age_EPIC # vector with patients' c-age (only EPIC patients)
betas_EPIC # matrix epic betas

```

# 2. Basic Epigenetic clocks

## 450K Batch

```{r }
cpgs.missing <- checkClocks(betas_450K)
```

Looks great.

## EPIC Batch

```{r }
cpgs.missing <- checkClocks(betas_EPIC)
```

Hannum' and Horvat's clocks have a relatively high proportion of missing CpGs, although acceptable (<20%). This is something expected in EPIC chip. 

# 2. B-Age calculation

## 2.1. 450K Batch

We now calculate the clocks and plot them against chrono-age.

**Hannum and TL clocks show an important number of missing cpgs**. BNN cannot be computed (>1 missing cpg horvath). 

```{R include=FALSE}
Bage_450K <- DNAmAge(betas_450K, 
                     age = c_age_450K, 
                     min.perc = 0.3, 
                     clocks = c('Horvath', 'Hannum', 'Levine', 'BNN',
                                'skinHorvath', 'PedBE', 'Wu', 'TL', 
                                'BLUP', 'EN'))
```

Performance of main clocks (we exclude skin and pediatric clocks):

```{r, fig.width=8, fig.height=5, fig.align='center'}
Bage_450K %>%
  select(age, Hannum, Horvath, Levine, BLUP, EN, TL) %>%
  tidyr::pivot_longer(cols = c('Hannum', 'Horvath', 'Levine', 'BLUP', 'EN', 'TL')) %>%
  ggplot(., aes(age, value)) +
  geom_point(alpha = .3) +
  geom_smooth(method = 'lm', se = F, color = 'firebrick') +
  facet_wrap(~name, scales = 'free') +
  ggpubr::stat_cor(method = 'pearson') +
  theme_classic() +
  theme(legend.position = 'bottom')
```

## 2.2. EPIC

Now repeat the same for EPIC

```{R include=FALSE}
Bage_EPIC <- DNAmAge(betas_EPIC, 
                     age = c_age_EPIC, 
                     min.perc = 0.3, 
                     clocks = c('Horvath', 'Hannum', 'Levine', 'BNN',
                                'skinHorvath', 'PedBE', 'Wu', 'TL',
                                'BLUP', 'EN'))
```

Performance of main clocks:

```{r, fig.width=8, fig.height=5, fig.align='center'}
Bage_EPIC %>%
  select(age, Hannum, Horvath, Levine, BLUP, EN, TL) %>%
  tidyr::pivot_longer(cols = c('Hannum', 'Horvath', 'Levine', 'BLUP', 'EN', 'TL')) %>%
  ggplot(., aes(age, value)) +
  geom_point(alpha = .3) +
  geom_smooth(method = 'lm', se = F, color = 'firebrick') +
  facet_wrap(~name, scales = 'free') +
  ggpubr::stat_cor(method = 'pearson') +
  theme_classic() 
```

# 3. EPITOC

Finally, we'll calculate EPITOC: 

- https://rdrr.io/github/metamaden/cgageR/

- https://genomemedicine.biomedcentral.com/articles/10.1186/s13073-020-00752-3

We don't need thje cpg Name: 

```{r }
betas_EPIC$CpGName <- NULL
betas_450K$CpGName <- NULL

if(all(colnames(betas_450K) == Bage_450K$id)){
  message('Files are correctly arranged for 450K')
} else  stop('Arrange files for 450k')
if(all(colnames(betas_EPIC) == Bage_EPIC$id)) {
  message('Files are correctly arranged for epic')
} else stop('Arrange files for EPIC')
```

## 3.1. 450K

### CgageR

First method of calculation using *cgageR* library:

```{r }
library(cgageR)

Bage_450K$Epitoc <- getAgeR(betas_450K, epitoc = T)$EpiTOC.output[['EpiTOC.Est']]$EpiTOC.Est
```

```{r, fig.width=3, fig.height=3, fig.align='center'}
Bage_450K %>%
  ggplot(., aes(age, Epitoc)) +
  geom_point(alpha = .3) +
  geom_smooth(method = 'lm', se = F, color = 'firebrick') +
  ggpubr::stat_cor(method = 'pearson') +
  theme_classic()
```

We don't expect a high correlation between C-Age and B-age in this clock. Please, REVIEW the manuscript: https://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-1064-3 
**>>Supplemental file 3<<**

## 3.2. EPIC

### CGAGER

```{r }
library(cgageR)
Bage_EPIC$Epitoc <- getAgeR(betas_EPIC, epitoc = T)$EpiTOC.output[['EpiTOC.Est']]$EpiTOC.Est
```

```{r, fig.width=3, fig.height=3, fig.align='center'}
Bage_EPIC %>%
  ggplot(., aes(age, Epitoc)) +
  geom_point(alpha = .3) +
  geom_smooth(method = 'lm', se = F, color = 'firebrick') +
  ggpubr::stat_cor(method = 'pearson') +
  theme_classic()
```

# 4. Save!
